The Design and Implementation of the Synchronous Language \CEU

### 2016-09-19 paper.tex
...
(>) temporal analysis -> concurrency analysis

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
more fine-grained control over program execution.
%
\CEU uses an event-triggered notion of time that enables compile-time temporal 
analysis to detect conflicting statements, resulting in deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we present the particularities of our design, such as 
stack-based internal events, the temporal analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU has a simple semantics with fine-grained control for intra-reaction 
execution, and is amenable to a temporal analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static temporal analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C, which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

In the subsections that follow, we discuss the main differences from \CEU to 
Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static temporal analysis for concurrent statements
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish the section with a summary of our design
    (\ref{sec.ceu.summary}).
%
We present the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we consider that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
though \code{A} and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to illustrate how to detect simultaneous 
button clicks in \CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events in \CEU never overlap because they are atomic.
Automatic mutual exclusion simplifies the reasoning about concurrency and is a 
prerequisite for the temporal analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

\subsection{ Static Temporal Analysis for Concurrent Statements }
\label{sec.ceu.analysis}

@@

To mitigate this threat, \CEU performs a temporal analysis at compile time to 
detect concurrent accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} (\ax:2,5)
      \textbf{cannot} be concurrent because they are \textbf{not} in parallel trails.
\item The assignments to \code{x} (\ax:5,8)
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} (\bx:5,8)
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
that precede a variable access and keep a list with all corresponding awaking 
events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

@@

\subsubsection{Temporal Analysis}

As a safety measure, the temporal analysis of Section~\ref{sec.ceu.analysis} 
also considers concurrent calls and accesses to external symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block with standard declarations in \C (ln.  
1--6).
%
During the boot reaction, two trails react concurrently inside the 
\code{par/and} (ln. 7--11):
the first trail calls symbol \code{\_f} (ln. 8),
while the second calls \code{\_g} and \code{\_id}, and also reads \code{\_NUM} 
(ln. 10).
%
Since \CEU does not inspect any code in \C, it complains about suspicious 
concurrent accesses between \code{\_f} and all symbols in the second trail.

@@

\CEU aims to offer a simpler semantics than Esterel with more determinism and 
fine-grained control over program execution.
%
The following list summarizes the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item \emph{Event-triggered notion of time bound to the semantics of the 
language.}
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with dealing with events in 
isolation, which simplifies the reasoning about concurrency.
In addition, the uniqueness of external events is a prerequisite for the 
temporal analysis of \CEU.
%
\item \emph{Deterministic intra-reaction execution and communication.}
Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
levels of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
stack-based internal events.
Programmers can always figure out which statement executes next, making runtime 
analysis and debugging easier.
%
\item \emph{Static temporal analysis for concurrent statements.}
Although execution is deterministic, the \CEU compiler still advises about 
suspicious statements that can react concurrently to the same event. 
%appropriate serialization.
%
\item \emph{Safe integration with \C.}
When dealing with concurrent system calls, programmers can provide annotations 
to improve the temporal analysis results, or to force non commutative 
concurrent behavior.
\CEU also requires finalization clauses to handle pointers representing 
resources.
%
\item \emph{First-class timers with dedicated syntax and automatic 
synchronization.}
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of adjusting timers in sequence and in parallel.
%
\end{itemize}

@@

The compilation process of programs in \CEU is composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{temporal analysis phase} detects inconsistencies in programs, such as 
unbounded loops and suspicious concurrent statements;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and packs it with platform-dependent functionality (e.g., system calls) and the 
runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.

@@

In the subsections that follow, we discuss implementation details specific to 
\CEU:
temporal analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

\subsection{Temporal Analysis for Concurrent Statements}
\label{sec.impl.analysis}

The compile-time temporal analysis phase detects inconsistencies in \CEU 
programs.
Here, we focus on the algorithm that detects suspicious concurrent statements, 
such as accesses to shared variables, as discussed in 
Section~\ref{sec.ceu.analysis}.

@@

Note that, as discussed in Section~\ref{sec.ceu.analysis}, the temporal 
analysis of \CEU infers precisely trails that are concurrent and yet do not 
share resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with multi-core execution are actually slower than 
single-core execution in the same system.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising and also appears in related 
work~\cite{esterel.multicore.1,esterel.multicore.2}.
One reason is the overhead from continuous fork-and-rejoin in small reactions.
Another reason is contention from excessive locality of data in stackless 
trails sharing contiguous static memory.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the temporal 
analysis that enables safe shared-memory concurrency.

Another unique aspect of \CEU is the distinction between external and internal 
events.
Internal events support stack-based micro reactions within external reactions, 
providing more fine-grained control for intra-reaction execution.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU goes one step further and performs a temporal analysis to detect trails 
that, when reordered, change the observable behavior of the program, i.e., 
trails that actually rely on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the temporal analysis of \CEU.
%
\CEU also guarantees deterministic behavior for timer compositions by adjusting 
inaccuracies in the system clock.

>>>>

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
more fine-grained control over program execution.
%
\CEU uses an event-triggered notion of time that enables compile-time concurrency 
analysis to detect conflicting statements, resulting in deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we present the particularities of our design, such as 
stack-based internal events, the concurrency analysis, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU has a simple semantics with fine-grained control for intra-reaction 
execution, and is amenable to a concurrency analysis that
improves safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency analysis to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.analysis}).
%
\item Safe integration with \C, which enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of our simple semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

In the subsections that follow, we discuss the main differences from \CEU to 
Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static concurrency analysis
    (\ref{sec.ceu.analysis});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish the section with a summary of our design
    (\ref{sec.ceu.summary}).
%
We present the formal specification of the semantics of \CEU in a separate 
work~\cite{ceu.phd}.

@@

We decided for the unique and queue-based semantics in \CEU for the reasons 
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is too abstract and imprecise:}
Outside the domain of hardware specification, a tick has no natural counterpart 
in the real world.
Also, since ticks require no time regularity~\cite{esterel.multiclock}, the two 
approaches for Esterel in Figure~\ref{fig.ticks} are legitimate, but lead to 
different behaviors for the same sequence of inputs.
%
\item \textbf{Events are never absolutely simultaneous:}
From a rigorous point of view, event occurrences are infinitesimal, having zero 
probability of being simultaneous.
This way, we consider that the notion of simultaneity should not be imposed by 
the language, but defined explicitly for each use case.
In the case of Esterel, simultaneity depends on the imprecise length of 
discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
Esterel considers the events \code{B} and \code{C} to be simultaneous, even 
though \code{A} and \code{B} ``physically'' happen much closer to one another.
In Section~\ref{sec.ceu.timers}, after introducing internal events and 
first-class timers, we show an example to illustrate how to detect simultaneous 
button clicks in \CEU.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to multiple events in \CEU never overlap because they are atomic.
Automatic mutual exclusion simplifies the reasoning about concurrency and is a 
prerequisite for the concurrency analysis to be discussed in 
Section~\ref{sec.ceu.analysis}.
\end{itemize}

@@

\subsection{ Static Concurrency Analysis }
\label{sec.ceu.analysis}

@@

To mitigate this threat, \CEU performs a concurrency analysis at compile time to 
detect concurrent accesses to shared variables:
if a variable is written in a trail segment, then a concurrent trail segment 
cannot read or write to that variable, nor dereference a pointer of that 
variable type.
%An analogous policy is applied for pointers \emph{vs} variables and pointers 
%\emph{vs} pointers.
%
Concurrency in \CEU is characterized when two or more trail segments in 
parallel react to the same input event.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
Considering the examples in Figure~\ref{lst.shared}:
%
\begin{itemize}
\item The assignments to \code{x} (\ax:2,5)
      \textbf{cannot} be concurrent because they are \textbf{not} in parallel trails.
\item The assignments to \code{x} (\ax:5,8)
      \textbf{cannot} be concurrent because they \textbf{cannot} execute during 
      the same reaction.
\item The assignments to \code{y} (\bx:5,8)
      \textbf{can} be concurrent because they are in parallel trails and 
      \textbf{can} execute during the same reaction.
\end{itemize}
%
The algorithm for the analysis, which is depicted in 
Section~\ref{sec.impl.analysis}, inspects all possible \code{await} statements 
that precede a variable access and keep a list with all corresponding awaking 
events.
Then, it checks all accesses in parallel trails to see if they share an awaking 
event.
If it is the case, the compiler warns about the suspicious accesses.

@@

\subsubsection{Concurrency Analysis}

As a safety measure, the concurrency analysis of Section~\ref{sec.ceu.analysis} 
also considers concurrent calls and accesses to external symbols in \C.
%
As an example, the program in Figure~\ref{lst.c}.a defines four external 
symbols inside a \code{native} block with standard declarations in \C (ln.  
1--6).
%
During the boot reaction, two trails react concurrently inside the 
\code{par/and} (ln. 7--11):
the first trail calls symbol \code{\_f} (ln. 8),
while the second calls \code{\_g} and \code{\_id}, and also reads \code{\_NUM} 
(ln. 10).
%
Since \CEU does not inspect any code in \C, it complains about suspicious 
concurrent accesses between \code{\_f} and all symbols in the second trail.

@@

\CEU aims to offer a simpler semantics than Esterel with more determinism and 
fine-grained control over program execution.
%
The following list summarizes the contributions of our design in this 
direction:
%
\begin{itemize}
%
\item \emph{Event-triggered notion of time bound to the semantics of the 
language.}
Event-driven programming is popular in many domains, such as server and GUI 
development.
We believe that programmers are more familiar with dealing with events in 
isolation, which simplifies the reasoning about concurrency.
In addition, the uniqueness of external events is a prerequisite for the 
concurrency analysis of \CEU.
%
\item \emph{Deterministic intra-reaction execution and communication.}
Determinism in \CEU is ``all-inclusive'' and does not depend on additional 
levels of static analysis.
It encompasses the whole language, including memory accesses, system calls, and 
stack-based internal events.
Programmers can always figure out which statement executes next, making runtime 
analysis and debugging easier.
%
\item \emph{Static concurrency analysis.}
Although execution is deterministic, the \CEU compiler still advises about 
suspicious statements that can react concurrently to the same event. 
%appropriate serialization.
%
\item \emph{Safe integration with \C.}
When dealing with concurrent system calls, programmers can provide annotations 
to improve the static analysis results, or to force non commutative 
concurrent behavior.
\CEU also requires finalization clauses to handle pointers representing 
resources.
%
\item \emph{First-class timers with dedicated syntax and automatic 
synchronization.}
Given the omnipresence of timers in embedded systems, a dedicated syntax can 
simplify the development and readability of programs.
Furthermore, automatic synchronization releases the programmer from the burden 
of adjusting timers in sequence and in parallel.
%
\end{itemize}

@@

The compilation process of programs in \CEU is composed of three main phases:
the \emph{parsing phase} converts the source code in \CEU to an abstract syntax 
tree (\emph{AST});
the \emph{concurrency analysis phase} detects inconsistencies in programs, such as 
unbounded loops and suspicious concurrent statements;
the \emph{code generation phase} converts the \emph{AST} to standard \C code 
and packs it with platform-dependent functionality (e.g., system calls) and the 
runtime of \CEU, compiling everything with \emph{gcc} to generate the final 
binary.

@@

In the subsections that follow, we discuss implementation details specific to 
\CEU:
concurrency analysis for determinism (Section~\ref{sec.impl.analysis}),
static memory allocation for data and trails (Sections~\ref{sec.impl.data} 
and~\ref{sec.impl.trails}),
static scheduling and trail finalization (Section~\ref{sec.impl.scheduler}),
single-threaded dispatching (Section~\ref{sec.impl.single}),
interaction with the environment (Section~\ref{sec.impl.environment}),
and the \VM back end (Section~\ref{sec.terra}),

\subsection{Concurrency Analysis}
\label{sec.impl.analysis}

The compile-time concurrency analysis phase detects inconsistencies in \CEU 
programs.
Here, we focus on the algorithm that detects suspicious concurrent statements, 
such as accesses to shared variables, as discussed in 
Section~\ref{sec.ceu.analysis}.

@@

Note that, as discussed in Section~\ref{sec.ceu.analysis}, the concurrency 
analysis of \CEU infers precisely trails that are concurrent and yet do not 
share resources.
Hence, these non-conflicting trails could potentially execute with real 
parallelism in multiple cores.
%
However, our experiments with multi-core execution are actually slower than 
single-core execution in the same system.
Considering that we use \CEU primarily in control-dominated applications, this 
result is not surprising and also appears in related 
work~\cite{esterel.multicore.1,esterel.multicore.2}.
One reason is the overhead from continuous fork-and-rejoin in small reactions.
Another reason is contention from excessive locality of data in stackless 
trails sharing contiguous static memory.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
We consider that this approach is more familiar to programmers and simplifies 
the reasoning about concurrency.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the concurrency 
analysis that enables safe shared-memory concurrency.
%
Another unique aspect of \CEU is the distinction between external and internal 
events.
Internal events support stack-based micro reactions within external reactions, 
providing more fine-grained control for intra-reaction execution.

Like \CEU, many synchronous languages rely on deterministic scheduling to 
preserve intra-reaction determinism (%
\emph{Reactive C}~\cite{rp.rc},                 % 1991
\emph{Protothreads}~\cite{wsn.protothreads},    % 2006
\emph{SOL}~\cite{wsn.sol},                      % 2007
\emph{SC}~\cite{rp.synchc},                     % 2009
and
\emph{PRET-C}~\cite{rp.pretc}).                 % 2010
%
%It is the responsibility of the programmer to specify the execution order for 
%threads, based on either explicit priorities, or source code lexical order 
%(similar to \CEU).
%
\CEU goes one step further and performs a concurrency analysis to detect trails 
that, when reordered, change the observable behavior of the program, i.e., 
trails that actually rely on deterministic scheduling.
%
%These languages have a tick-based notion of time similar to Esterel, which 
%prevents the concurrency analysis of \CEU.
%
\CEU also guarantees deterministic behavior for timer compositions by adjusting 
inaccuracies in the system clock.
